Water desalination using polyelectrolyte hydrogel. Gibbs ensemble modelling

### 2022-06-15 gibbs.tex
lyx
Switched to tex

\documentclass{elsarticle}
\input{definitions.tex}
\usepackage[margin=2.5cm]{geometry}% by courtesy of Mico
\begin{document}
\begin{frontmatter}{}

\title{\mytitle}

%\title{The \PE{} hydrogels as draw agents for desalination of solutions with multivalent ions}

%\tnotetext[t1]{This document is a collaborative effort.}
%\tnotetext[t2]{The second title footnote which is a longer longer than the first one and with an intention to fill in up more than one line while formatting.}

    \author[cuni,imc]{Michail Laktionov}
    \ead{michail.laktionov@natur.cuni.cz}
    \ead[url]{http://www.physchem.cz/}
    \author[cuni]{Oleg V. Rud}
    \author[cuni]{Lucie~Nova}
    %\author[cuni]{Peter~Kosovan}
    \author[cuni]{Filip~Uhlik}

    \address[cuni]{Department of Physical and Macromolecular Chemistry, Faculty of Science, Charles University in Prague, Hlavova 8, Praha 2 128 00, Czech Republic}
    \address[imc]{Institute of Macromolecular Compounds of Russian Academy of Sciences, 199004, Bolshoy pr. 31, Saint-Petersburg, Russia}

\begin{abstract}
By this article we model compression of \PE{} hydrogel in equilibrium with small volume of salty aqueous solution.
We show that the the change of the hydrogel volume affects the salinity of the solution. The decrease of the gel volume decreases the solution and vice versa the swelling of the gel increases the surrounding solution. 
This effect can be employed for water desalination ...
\end{abstract}

\end{frontmatter}{}

@@

\reffig{fig:diamond} presents a diamond-like polymer network inserted in a simple cubic simulation box of volume $V_{\mathrm{box}}$ with periodic boundaries.
The network consists of $16$ linear \PE{} chains, each chain comprising $30$ monomer units.
Each unit is able to change its ionization state from charged to neutral and vice versa according to the ionization equilibrium.
The mobile ions ($\na$, $\cl$ and $\ca$) are present in certain amount such that they fulfill the electroneutrality condition.

We model the network in equilibrium with the bulk of aqueous solution.
In order to keep the equilibrium conditions we consider three processes to occur:

\begin{enumerate}
	\item the thermal movement of all segments in simulation box (Langevin dynamics);
	\item the transfer of ions between two vessels:
        \item the change of ratio between volumes of the two vessels emulationg the piston applying certain pressure to one of the vessels:
%\todoi{We shall get rid of $\so$ ions and mention it in the inroduction. The reason is that their concentration is small, and they do not enter the hydrogel }
\end{enumerate}

All three processes occur simultaneously in real systems, however, we simulate them alternating each other in stop-run mode in a random order.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\columnwidth]{{figures/vmdscene}.pdf}
    \caption{
       	Diamond-like network in the simulation box. 
	Color code represents the individual ion types
	(red: $\na$, blue: $\cl$, yellow: $\ca$) and
        the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment ($\A$)).
    }
    \label{fig:diamond}
    %in order to make snapshot use file snapshot.py
    % in order to edit the existing snapshot use file data/gel_MPC30_box_l53.0_pK7.00_pH7.00_pCl2.00_pCa2.94_lB0.0.vmd
\end{figure}
\todo{remove $\ca$ ions from snapshot}
\subsection{Langevin dynamics\label{MD}}

We use three types of interactions, \ie bonded interactions, short-range non-bonded interactions and electrostatic interactions.

We use Finite extension nonlinear elastic (FENE) potential~\refeq{eq:fene} to account for the bonding interactions.
%The potential accounts for nonlinear elastic extensions and as a consequence, the simulation is not constrained too much.
\begin{equation}\label{eq:fene}
    V_\mathrm{FENE}(r) = -\frac{1}{2} K \Delta r_\mathrm{max}^2\ln \left[ 1 - \left(\frac{r-r_0}{\Delta r_\mathrm{max}} \right)^2 \right ],
\end{equation}
where $K$ is the magnitude of symmetric interaction between two segments,
$\Delta r_{\mathrm{max}}$ is the maximal stretching length of the bond and $r_0$ is the equilibrium bond length.
In our simulations we set $K = 10 \kT/\sigma^{2}$, $\Delta r_{\mathrm{max}} = 2 \sigma$ and $r_0 = 1.0\sigma$~\cite{Jin2007}.

We model the non-bonded interactions using truncated Lennard-Jones interaction potential.
This potential imposes strong repulsion between all particles at short distances:
\begin{equation}
    \begin{split}\label{eq:LJ}
        V_\mathrm{LJ}(r) =
        \begin{cases}
            4 \varepsilon \left( \left(\cfrac{\sigma}{r}\right)^{12}
            - \left(\cfrac{\sigma}{r}\right)^6\right)
            & \mathrm{if~} r < r_\mathrm{cut}\\
            0
            & \mathrm{elsewhere}
        \end{cases},
    \end{split}
\end{equation}
where $r$ is the interparticle distance,
$\sigma = 0.35 \mathrm{nm}$ is the characteristic size of the particles,
$\varepsilon$ is the depth of the potential well in $\kT$ units and $r_\mathrm{cut}$ is the cut-off distance beyond which the potential is set zero.

We model the long range electrostatic interactions by the Coulomb potential:
\begin{equation}
    V_{\mathrm{EL}}=\lb\kT \cfrac{q_{1}q_{2}}{r},
\end{equation}
where $\lb$ is Bjerrum length. We set $\lb = 2\sigma = 0.7 \mathrm{nm}$ which corresponds to water at $T=300 \mathrm{K}$.

We use Langevin thermostat, \ie
we add two additional terms to force in equation of motion
\begin{equation}
f_i =  -\gamma v_i(t)+\sqrt{2\gamma \kT }\eta_i(t),
\end{equation}
where the first term correspond to constant friction with $\gamma$ being a friction coefficient, and the second one corresponds to random thermal force with $\eta_i$ being a normally distributed random number (see for details \cite{Grest1986}).

%where $\gamma$ is the friction coefficient and .
%\todo{there is friction and there is random force ballanced by fluctiation dissipation theorem so that it is stationary}
%which causes the decrease of the particles' velocities.
%This is implemented by addition of two corresponding terms to the particles' equations of motion, namely
%\todo{usually -gamma v to keep gamma>0}
%where the first term correspond to

%\paragraph{Units} % apart from cref reference all references are in the text. Do we need cref?
%By setting up the interaction potentials we defined the system of units.
%We have chosen the unit of length being based on the Bjerrum length in water at room temperature, $l_B \simeq 7$nm, and $\sigma=0.5l_B = 0.35$.
% % %Then the unit of concentration is $\cref=(\sigma^{3}\NA1000)^{-1}$mol/l. % % % do we need to say that?
%We set the energy unit to be $\varepsilon = \kT$.

\section{Sampling in the Gibbs ensemble\label{sec: sampling the gibbs ensemble}}
All thermal moves are frozen during the sampling of the ion pairs exchange.
The system represents the ensemble with constant number of particles distributed between two vessels (I) and (II).

In the following paragraph we derive the criteria for sampling of the ion exchange equilibrium.
\paragraph{Gibbs canonical ensemble}
Let us first consider the total free energy of both vesselsassuming that we have only particles of a single type:
\begin{equation}
    F=\left(E_{I}-TS_{I}\right) + \left(E_{II}-TS_{II}\right) \label{eq:F-gibbs}
\end{equation}
where $E$ is internal energy, $T$~--- temperature, $S$~--- entropy.

%Our model system represents a finite volume $V$ which exchanges particles with the bath of infinite volume.
%The infinity of the volume means, that the particle exchange between the system and the bath does not influence on the chemical potentials in the bath.
%Such ensemble is called grand canonical ensemble; the free energy in this ensemble, the Landau potential, is
The entropy $S$ is given by the Boltzmann formula~\cite{Nagle2004}:
\begin{equation}
    S=\kB\ln\frac{V^{N}}{N!}\label{eq:entropy}
\end{equation}
%\todo{for partiles without interactions}
where $N$ is a number of particles in corresponding vessel.
This formula accounts for two contributions:
\begin{enumerate}
    \item the combinatorial entropy $S_{c}=-\kB\ln N!$ which reflects the freedom of choice among $N$ particles
    \item the mixing entropy $S_{m}=\kB N\ln V$ which reflects the freedom to place the chosen particle randomly within the simulation box. 
\end{enumerate}
$V$ is the unitless volume, \ie the volume measured in units of $\sigma^3$. 
Thus \refeq{eq:F-gibbs} turns to 
%\todo{this is not reason, moreover A.10 is already wrong as for units}
\begin{equation}
    F=\EI-\kT\ln\frac{\VI^{\NI}}{\NI!}\ + \ \EII-\kT\ln\frac{\VII ^{\NII}}{{\NII} !}
\label{eq: F_gibbs}
\end{equation}
When a single particle transferred from the vessel (I) to the  vessel (II) the free energy in vessel (I) changes 
\begin{equation}
\Delta \FI = \Delta \EI+\kT\ln\frac{\VI^{\NI}}{\NI !} - \kT\ln\frac{\VI^{\NI-1}}{(\NI-1)!}=\Delta \EI + \kT\ln\frac{\VI}{\NI}
\end{equation}
whwereas the free energy change in (II) vessel is
\begin{equation}
\Delta \FII = \Delta \EII+\kT\ln\frac{\VII^{\NII}}{\NII !} - \kT\ln\frac{\VII^{\NII+1}}{(\NII+1)!}=\Delta \EII + \kT\ln\frac{\NII+1}{\VII}
\end{equation}
Thus the total change of free energy will be 
\begin{equation}
\Delta F = \Delta \EI+\Delta \EII+\kT\ln\frac{\VI}{\VII}\frac{\NII+1}{\NI}
\end{equation}
Introducing $\xi=\pm 1$, which reflects the direction of movement of particle, $\xi$= +1 if the particles moves from vessel (I) to vessel (II), and $\xi=-1$ if the particles moves backwards, then the previous formula transfers to

\begin{equation}
\Delta F = \Delta E_{tot}+\kT\xi\ln\frac{\VI}{\VII}\frac{\NII+\theta(\xi)}{\NI+\theta(-\xi)}
\end{equation}
where $\theta(x)$ is the Heaviside function.

>>>>

 %% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
%\documentclass{achemso}
\documentclass[journal=mamobx, layout=twocolumns,manuscript=article]{achemso}

\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{verbose}
\usepackage{xcolor}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=false,
 breaklinks=true,pdfborder={0 0 0},pdfborderstyle={},backref=false,colorlinks=true]
 {hyperref}
\usepackage[textsize=small]{todonotes}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.

\title{Water desalination using polyelectrolyte hydrogel. Gibbs ensemble
modelling}

\author{Mikhail Laktionov}

\affiliation[cuni]{Department of Physical and Macromolecular Chemistry, Faculty of Science,
Charles University in Prague, Czech Republic}

\alsoaffiliation[itmo]{St. Petersburg National Research University of Information Technologies,
Mechanics and Optics, St. Petersburg, Russia.}

\author{Lucie Nová}
\author{Oleg V. Rud}

\affiliation{Department of Physical and Macromolecular Chemistry, Faculty of Science,
Charles University in Prague, Czech Republic}

\alsoaffiliation[imc]{Institute of Macromolecular Compounds of Russian Academy of Sciences,
Saint-Petersburg, Russia}

\email{oleg.rud@natur.cuni.cz}

\abbreviations{MC, MD, RO, FO}

\keywords{polyelectrolye hydrogel, simulation, desalination}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
%\usepackage[utf8]{inputenc}
\usepackage{amssymb}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\newcommand{\ie}{\textit{i.~e.} }
\newcommand{\cref}{c^{\ominus}}
%\newcommand{\cs}{c_{s}}
\newcommand{\kT}{k_\mathrm{B}T}
\newcommand{\kB}{k_\mathrm{B}}
\newcommand{\lb}{l_\mathrm{B}}
\newcommand{\NA}{N_{\mathrm{A^-}}}
\newcommand{\muna}{\mu_\mathrm{Na^+}}

\newcommand{\mucl}{\mu_\mathrm{Cl^-}}
\newcommand{\muca}{\mu_\mathrm{Ca^{2+}}}
\newcommand{\muh}{\mu_\mathrm{H^+}}
\newcommand{\mua}{\mu_\mathrm{A^-}}
\newcommand{\muha}{\mu_\mathrm{HA}}
\newcommand{\muoh}{\mu_\mathrm{OH}}

\newcommand{\cna}{c_\mathrm{Na^+}}
\newcommand{\ccl}{c_\mathrm{Cl^-}}
\newcommand{\cca}{c_\mathrm{Ca^{2+}}}
\newcommand{\ch}{c_\mathrm{H^+}}
\newcommand{\cp}{c_\mathrm{p}}
\newcommand{\nna}{N_\mathrm{Na^+}}
\newcommand{\ncl}{N_\mathrm{Cl^-}}
\newcommand{\ncleq}{\widetilde{N}_\mathrm{Cl^-}}
\newcommand{\nca}{N_\mathrm{Ca^{2+}}}

\newcommand{\superin}{^\mathrm{in}}
\newcommand{\subin}{_\mathrm{in}}
\newcommand{\subi}{_\mathrm{i}}

\newcommand{\gel}{^\mathrm{gel}}
\newcommand{\tot}{^\mathrm{tot}}
\newcommand{\out}{^{out}}
\newcommand{\coh}{c_\mathrm{OH}}
\newcommand{\bulk}{^{\mathrm{b}}}
\renewcommand{\H}{\mathrm{H^+}}
\newcommand{\A}{\mathrm{A^-}}
\newcommand{\AH}{\mathrm{AH}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}
\newcommand{\todoi}[1]{\todo[inline]{#1}}
\setuptodonotes{fancyline, color=blue!30, size=\tiny}

\newcommand{\cl}{\mathrm{Cl^-}}
\newcommand{\br}{\mathrm{Br^-}}
\newcommand{\na}{\mathrm{Na^+}}
\newcommand{\h}{\mathrm{H^+}}
\newcommand{\ka}{\mathrm{K^+}}
\newcommand{\oh}{\mathrm{OH^-}}
\newcommand{\ca}{\mathrm{Ca^{2+}}}
\newcommand{\mg}{\mathrm{Mg^{2+}}}
\newcommand{\so}{\mathrm{SO_4^{2-}}}

\newcommand{\EI}{E_{\mathrm{I}}}
\newcommand{\EII}{E_{\mathrm{II}}}
\newcommand{\SI}{S_{\mathrm{I}}}
\newcommand{\SII}{S_{\mathrm{II}}}
\newcommand{\NI}{N_{\mathrm{I}}}
\newcommand{\NII}{N_{\mathrm{II}}}
\newcommand{\VI}{V_{\mathrm{I}}}
\newcommand{\VII}{V_{\mathrm{II}}}
\newcommand{\FI}{F_{\mathrm{I}}}
\newcommand{\FII}{F_{\mathrm{II}}}
\newcommand{\nnaI}{N^{\na}_{\mathrm{I}}}
\newcommand{\nclI}{N^{\cl}_{\mathrm{I}}}
\newcommand{\nnaII}{N^{\na}_{\mathrm{II}}}
\newcommand{\nclII}{N^{\cl}_{\mathrm{II}}}

\newcommand{\Ka}{K_{\mathrm{A}}}
\newcommand{\pKa}{\mathrm{p}\Ka}
\newcommand{\pK}{\mathrm{p}K}
\newcommand{\pH}{\mathrm{pH}}
\newcommand{\mol}{\mathrm{mol}}
\newcommand{\molperl}{\mathrm{mol/l}}
\newcommand{\kg}{\mathrm{kg}}
\newcommand{\res}{^{\mathrm{res}}}
\newcommand{\pHres}{\pH\res}
\newcommand{\pHgel}{\pH\gel}
\newcommand{\cs}{c_{\mathrm{s}}}
%\newcommand{\cs}{c_{\mathrm{salt}}}
\newcommand{\csres}{\cs\res}
\newcommand{\Vgel}{V\gel}
\newcommand{\Vgeleq}{\widetilde{V}\gel}
\newcommand{\Pgel}{P\gel}
\newcommand{\Pres}{P\res}
\newcommand{\Vout}{V\out}
\newcommand{\Vbox}{V\tot}
\newcommand{\PE}{polyelectrolyte{}}

\newcommand{\reffig}[1]{Figure~\ref{#1}}
\newcommand{\refeq}[1]{Equation~\ref{#1}{}}
\usepackage{afterpage}
\newcommand\blankpage{%
    \null
    \thispagestyle{empty}%
    \addtocounter{page}{-1}%
    \newpage}

\newcommand{\mytitle}{Gibbs ensemble for \PE{} hydrogel}
\newcommand{\etal}{\textit{et al.}{}}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}
\makeatother

\begin{document}
\begin{abstract}
Recently polyelectrolyte hydrogels have been proposed as draw agents
for reverse osmosis desalination techniques. Indeed, polyelectrolyte
hydrogels have the ability to absorb a big amount of water across
forward osmosis membrane as a result of their swelling pressure. The
insoluble cross-linked network of the gel enables dewatering under
the influence of a stimuli (thermal and/or mechanical). On the other
hand, the network structure of a polymer hydrogel from a thermodynamic
perspective is already an osmotic membrane. So hydrogel microparticles
may allow to completely avoid the osmotic membranes in forward osmosis
and use microfiltration instead. By this article, we present our recent
theoretical study of the use of polyelectrolyte hydrogel for water
desalination. We modeled the thermodynamic equilibrium of coexistence
of the gel and the aqueous salt solution in the so-called closed ensemble,
in which the total amount of ions is assumed to be constant. We modeled
the compression of the gel and the associated with that release of
the solution. We have shown that the squeezed out solution has a little
lower salinity than that the gel was equilibrated with. Also, we performed
a set of simulations modeling the process of continuous decrease of
water salinity up to freshwater concentrations.

\end{abstract}

\section{Introduction\label{sec: intro}}

\paragraph{Water desalination technologies.}

Two basic approaches for separating water from salt are present in
modern desalination technology \cite{Miller2003,Curto2021}.

The first approach~--- distillation. It uses thermal means to effect
a phase change of the water to vapor (or solid), then to physically
separate the vapor phase from the remaining salt solution, and then
recover the thermal energy for reuse as the separated vapor reverts
to liquid form. Distillation processes were the first desalination
techniques conducted on a large commercial scale and still account
for a large portion of the modern world’s desalination capacity.

The second approach is to physically separate the brine components,
generally with a membrane, as they move in response to an externally
applied gradient. In the context of our study, we will mention the
reverse osmosis process (RO) \textasciitilde -{}-{}- the major processes
of all the modern desalination industry, and the new emerging membrane
technology -{}-{}- forward osmosis (FO) \cite{Akther2015}. In RO,
water passes through a membrane that is impermeable to the solute
in response to a chemical potential gradient achieved through pressurization.
In FO, a concentrated draw solution, with a lower water chemical potential,
and a feed solution, with a higher water chemical potential, are separated
by a membrane that rejects the salt but allows the passage of water.
The water permeation from feed solution to draw solution is a spontaneous
process driven by the chemical potential gradient.

Distillation is easy and cheap technology but it is characterized
by relatively high energy costs due to the dissipation of thermal
energy. On the contrary, RO uses expensive osmotic membranes, which
need to be replaced regularly because of scaling and fouling. Moreover,
RO requires very high operating pressures, ranging from 20 to 200
bar, to let the water pass through the membrane. But in terms of energy
losses, it works close to the thermodynamic limit. Thus, theoretically,
per one transferred from freshwater to salty water ion pair, it consumes
only the energy of difference of those ions chemical potentials.

The absence of large hydraulic pressure in FO process in comparison
to RO allows to reduce the energy consumption in pumping, reduce membrane
scaling and fouling, therefore significantly increase the lifetime
of the membranes. In FO, the draw solutes (agents) dispersed and/or
dissolved in water form homogeneous draw solutions. The correct choice
of the draw agents is a task of paramount importance. As an osmotically
driven process, the draw solute is expected to significantly reduce
the water chemical potential, and consequently generate a high osmotic
pressure. On the other hand, the diluted draw solution is also expected
to be easily separated to fresh water and to concentrated draw solution
for reuse \cite{Cai2016}.

\paragraph{Hydrogels for desalination.}

Hydrogels are three-dimensional networks of polymer chains that are
crosslinked by either physical or chemical bonds. They are able to
entrap large volumes of water attracted by the high concentration
of hydrophilic groups. Hydrogels with ionic groups on the comonomer
unit are able to reject salt ions from the solution, they absorb a
solution of lower salinity than that they are equilibrated with. When
a dehydrated or deswollen hydrogel uptakes water, its polymer chains
extend, creating a swelling pressure. For example, as reported in
\cite{Wack_2009} weakly crosslinked poly(acrylic acid)/poly(sodium
acrylate) copolymers with polymer volume fractions between 0.03 and
0.30 exhibit a swelling pressure ranging from 0.20--4.23 MPa.

An important, advantageous aspect of polymer hydrogels is that they
can undergo reversible volume change, \ie solution --- gel phase
transitions in response to environmental stimuli. This aspect causes
hydrogels to be labeled as ‘smart’ hydrogels. Many physical and chemical
stimuli have been applied to induce various responses of such smart
hydrogels, in particular, to change them from hydrophilic to hydrophobic,
thereby releasing water. These stimuli include: temperature, solvent
composition, light, mechanical pressure, sound, electric and/or magnetic
field, whilst the chemical or biochemical stimuli include pH, ions,
and specific molecular recognition events \cite{Tanaka_1982,Serizawa_2001,Lietor_Santos_2009,Qiu_2001}.

The smart properties of hydrogels were utilized for desalination in
a recent study by Li \etal \cite{Li2011}, where hydrogels have been
considered as draw agents for FO. The main point of the cited article
is that hydrogels, due to their swelling and osmotic pressure, were
able to absorb water across the FO membrane. The insoluble cross-linked
network of the gel enables dewatering under the influence of stimuli
(thermal and/or mechanical). For instance, Li \etal proposed the
use of hydrogel made of thermoresponsive polyelectrolyte~--- copolymer
of poly-N-isopropyl acrylamide (p-NIPAAm) and polyacrylic acid (p-AA).
Depending on the temperature the network of the gel may appear to
be either hydrophilic or hydrophobic. In the hydrophilic state, the
gel is used as a draw agent accumulating water inside, then in the
hydrophobic state, the water releases out.

Moreover the hydrogel can be as an osmotic membrane from a thermodynamics
perspective \cite{Wang_2014}. Such a view on hydrogels was employed
in a series of works by the group of prof. Wilhelm (see for example
\cite{Arens_2017,Fengler_2020}). The authors proposed to get rid of
the osmotic membrane and simply use only a micro filtration membrane
to compress the hydrogel and separate it from the solution. In their
method first, the deswollen hydrogel was equilibrated with saline
water feed. During equilibration, the gel was swelling absorbing water.
Then it was taken out from the feed solution and mechanically compressed
by means of a microfiltration membrane. The squeezed out brine was
obtained to have lower salinity than the feed water.

A similar approach was used by Ali \etal \cite{Ali2015}. In their
study the thermosensitive gel was used (based on copolymers p-NIPAAm
and p-AA), so instead of physical compression, the dewatering was
done by an external heating stimulus (sunlight). At night the gel
was equilibrating with feed water and at daytime, under sunlight,
the gel was shrinking and releasing the collected inside water.

\paragraph{Physics behind the desalination process.}

\begin{figure*}
\centering 
\includegraphics[width=0.5\textwidth]{figures/gc}\includegraphics[width=0.5\textwidth]{figures/gibbs}\caption{Diamond-like network in the simulation box. Color code represents
the individual ion types (red: $\na$, blue: $\cl$, yellow: $\ca$)
and the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment
($\A$)). \label{fig:open and closed}.}
\end{figure*}

The maximum entropy principle requires the density of charges to be
more-less the same throughout the system medium. Since the polyelectrolyte
gel has its own charges and its own neutralizing counterions, the
density of mobile ions (which are able to freely enter and leave the
gel) in gel appears to be smaller than their density outside the gel.
Therefore the internal solution of the gel has lower density of mobile
ions than the solution outside. In that sense, the gel acts as an
osmotic membrane separating solutions \cite{Rud2020}. The driving
force of the separation is the Donnan potential originated from the
charges of the hydrogel network. The density of mobile ions difference
between ions concentrations in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\[
\frac{c_{-}^{in}}{\cs}=\frac{\cs}{c_{+}^{in}}=\sqrt{1+\left(\frac{\alpha\cp}{2\cs}\right)^{2}}\pm\frac{\alpha\cp}{2\cs}
\]
where $c_{-}^{in}$ and $c_{+}^{in}$ are the internal concentration
of anions and cations, $\cs$ is the salinity of external solution
$\cp$ and $\alpha$ are the gel molar density and the ionization
degree. The $\pm$ sign in the formula depends on whether the gel
is made of polyanion or polycation.

The internal solution can be extracted by means of compression and/or
other stimulus providing that the charge of the gel remains constant.
In the case of gel made of weak (pH-sensitive) polyelectrolyte, the
compression initiates discharging of the gel, therefore the neutralizing
counterions may leave out the gel diminishing the desalination effect
and affecting the pH of the extracted solution. Nevertheless, the
release of counterions may be employed for desalination as well and
the way how to do it is discussed in\cite{Rud2018}. 
\begin{enumerate}
\item However, one can argue that sole Donnan effect is insufficient to
achieve a high salt rejection \cite{Cai2016} and the salinity of
the water squeezed from hydrogel under a very high hydraulic pressure
(up to 100 bar \cite{Fengler_2020}) turns to be not much different
from the initial.
\item Indeed, the use of high hydraulic pressure diminishes all the advantages
of this method over the reverse osmosis. At the same time the reversibility
of hydrogel swelling after such strong compression remains questionable. 
\item Nevertheless by the current study we would like to model the compression
of the gel limiting ourselves within low compression rates, less than
5 bar. We will show how the compression of the gel affects the surrounding
salinity and model the whole desalination process as a cascade of
$c$. 
\end{enumerate}
In the current study we would like to concentrate on the use of gel
made of strong polyelectrolyte for water desalination. Namely, to
investigate how the compression of the gel affects the salinity of
the surrounding the gel solution and to propose the way of water desalination
employing the hydrogel swelling/deswelling set of processes.

The target of our study is to model the thermodynamic equilibrium
between the two boxes and to show how the compression of the gel affects
the ion density in the outside solution.

@@

\begin{figure*}
\centering \includegraphics[width=0.5\textwidth]{figures/simgibbs}\caption{Diamond-like network in the simulation box. Color code represents
the individual ion types (red: $\na$, blue: $\cl$, yellow: $\ca$)
and the hydrogel (gray: neutral segment ($\AH$), cyan: charged segment
($\A$)). \label{fig:diamond}.}
\end{figure*}

\subsection{Model}

We model our gel as a network of 16 linear polymer chains, each by
30 monomer units. The chains are connected to a diamond-like polymer
network and put into a simulation box with periodic boundaries. Each
unit of the network carries a negative electric charge, which is equal
to the charge of electron. Except the particles of the network, the
mobile monovalent ions, $\na$ and $\cl$, are present in simulation
box. The total electric charge of all the particles in the box is
zero, therefore the amount of $\na$ ions is bigger than that of $\cl$
by the number of hydrogel units, $N\gel=16\cdot30=480$.

The goal of this study is to simulate the process of water desalination
by means of compression. In order to do it we simulate the gel in
two ensembles.
\begin{enumerate}
\item The first is so called \emph{open system}, when the simulation box
with gel freely exchanges ions with a big amount of aqueous solution
of certain constant salinity. 
\item And in \emph{closed system}, when the gel is in equilibrium with a
finite volume of aqueous solution.
\end{enumerate}
The compression of the gel in the open system does not affect the surrounding salinity, thus the chemical potential of ions remains constant $\muna=$ $\mucl=$ $\text{Const}$. 
On the contrary, what remains constant in the closed system is total number of ions, that are contained in the gel and in the external volume, $\nna=\text{Const}$ and $\ncl=\text{Const}$.
By $\nna$ and $\ncl$ we will understand the density of the corresponding ion in the compression volume
\begin{equation}
\ncl = \left(\ncl\gel + \cs\cdot(\Vbox - \Vgel)\right)/\Vbox
\end{equation}
\paragraph{Open system.}

We simulate the open system using the grand reaction method \cite{Landsgesel2020a,Rud2020}.
This method is a hybrid of molecular dynamics (MD) and Monte Carlo
(MC). The whole simulation represents a chain of subsimulations of
MD and MC followed one by each other. For MD simulation we used a
standard Langevin dynamics \cite{Grest1986} which models the mechanical
movement of all the system particles. Whereas MC simulates the thermodynamic
equilibrium with reservoir, which exchanges ions with the simulation
box. The insertion (and deletion) of ion pairs is considered as a
reaction of creation (or annihilation) of an ion pair 
\[
\varnothing\stackrel{K}{\leftrightarrows}\na+\cl
\]
with a reaction constant defined by the chemical potential of ions,
$K=\exp\left(\muna+\mucl\right).$

\paragraph{Closed system.}

The similar scheme we use to simulate the gel in closed system. The
procedure represents a chain of subsequent MD and MC simulation, but
now we run the molecular dynamics in two simulation boxes simultaneously.
The first box of the volume $\Vgel$ contains the gel with ions, and
the second box, $\Vout$--- contains only ions. These two boxes are
represented in Figure~\ref{fig:diamond}. The total volume of both
boxes is kept constant, $V=V_{\mathrm{gel}}+V_{\mathrm{out}}$, so
the compression of the gel implies the decrease $\Vgel$ and increase
$\Vout$.

The thermodynamic equilibrium between the two volumes implies the
exchange of the ion pairs between them, which is done by means of
MC procedure in a way similar to described in \cite{Panagiotopoulos1988b}.

\paragraph{Algorithm.}

In general the simulation algorithm is following:
\begin{enumerate}
\item simulate molecular dynamics in both boxes; collect the observables,
for examples pressure values $P$, to an array; 
\item perform Monte-Carlo procedure simulating ion pair exchange; collect
the arrays ofobservables sample : number of ions in both boxes, $\nna^{I}$,
$\ncl^{I}$, $\nna^{II}$ and $\ncl^{II}$ ; 
\item repeat the procedure until the desired length of sample arrays is
reached.
\end{enumerate}

\section{Results and discussion }

\paragraph{Compression in open system.}

First we run a set of simulations modeling the gel compression in \emph{open system}, \ie in equilibrium with big bath of certain salinity, $\cs$. 
The simulations are run for a set of different simulation box volumes. 
Each simulation returns the averaged pressure, $P$, and the number of $\cl$ ions present in the simulation box, $\ncl$.

In order to get the partial pressure of the gel, we substitute from $P$ the osmotic pressure of ions in reservoir: $P^{gel}=P-P^{res}$.
$P^{res}$, in turn, we calculate by running a separate simulation of a reservoir, containing ionic gas in equilibrium with the bath of the same salinity $\cs$. 
The gel partial pressure, $\Pgel$, is the pressure which needs to be applied to the gel via solvent permeable filter to compress the gel to a molar specific volume, $V$.

The dependencies of $\Pgel$ versus gel molar volume $\Vgel$ (the volume calculated per one mol of gel segment) are present in Figure \ref{fig: PV} for a set of different salinity. 
These dependencies corresponding to \emph{open system} are represented here by solid lines. 
For example, blue solid line illustrates the compression (or swelling) of the gel in equilibrium with a reservoir of salinity, $\cs=0.063$ mol/l. 
The points where the pressure equals to zero, $\Pgel=0$, (indicated by filled circles) are the gel \emph{free swelling equilibrium} states. 
It is seen, that the increase of salinity shifts the free swelling equilibrium towards smaller volumes, \ie the gel shrinks with increasing salinity. 
In general, all the solid curves shift towards smaller volumes with increasing salinity. 
This effect is well known, and is typical for all branched \emph{strong} polyelectrolytes. 
It is caused by the decrease of ions osmotic pressure and by screening of electrostatic interactions \cite{Zhulina2000,Landsgesel2020a}.\footnote{The salinity dependence of the size of \emph{weak} polyelectrolyte gel is in general non-monotonic \cite{Rud2018} but the case of weak polyelectrolyte is beyond the scope of this study. } %

\paragraph{Compression in closed system.}

\begin{figure*}[t]
\subfloat[Partial pressure of the gel versus gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub}}
\hspace{0.02\textwidth}
\subfloat[Supernatant salinity versus gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\todoi{replace $V$ by $\Vgel$ in Figures}
\end{figure*}

Each set of simulations in an \emph{open system} defines the free swelling equilibrium state of the gel for each salinity. 
Namely, each salinity correspond to a certain gel molar volume, and to a certain amount of ions in gel in equilibrium state, $\cs\longrightarrow$, $\left(\Vgeleq,\ \ncleq\right)$.
These equilibrium states we use as starting ones for the gel compression in \emph{closed system}. 
Thus, we simulate the compression of the gel in the closed volume, $\Vbox$, which is equal to $\Vgeleq$, and contains $\ncleq$ of ion pairs (apart from neutralizing counterions). 
So, we prepare two systems, one~--- simulating the gel of the volume $\Vgel$, and another~--- simulating the supernatant solution of the volume $\Vout = \Vbox - \Vgel$; in this way the total volume containing the gel and the supernatant solution remains constant.
Also, the ion pairs of the amount $\ncleq$ remain all the time constant and are shared between the two volumes. %For each different initial states we run a set of simulations emulating the compression of the gel, \ie varying simultaneously $\Vgel$ and $\Vout$. 
%At any all the ions are getting redistributed between the systems but their total amount remains constant and equal to $\ncl^{0}$.

The process of the gel compression in \emph{closed} system is depicted in Figure~\ref{fig: PV} (dotted lines). 
There the gel partial pressure is depicted as function of the gel molar volume.
In this plot, the blue dotted line, for example, illustrates the compression of the gel equilibrated with the solution of salinity $\cs=0.063$ mol/l, in a volume which gel has at zero pressure. 
Since, in this case, the volumes of the gel $\Vgel$ and of the supernatant solution $\Vout$ are comparable, the compression of the gel affects the salinity of supernatant solution. 
The salinity now decreases, changing from $\cs^{0}=0.063$ at free swelling equilibrium state to 0.045 mol/l at $P\gel=5$ bar.
